---
title: "Have Your Cake and Synthesize it Too"
subtitle: Synthesis Example
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Get and Prep Example Data

We'll use the `faketucky` dataset to provide an example of our "restricted" dataset.

_But first a quick rant about code style_. Notice the reference to the function: 

`packageName::functionName()`

_This is how you should write code if you want to avoid headaches in the future_, 
particularly when a function may exist in multiple packages.  Chasing down the 
package that a function comes from is not always the easiest/best use of time, 
but a few extra keystrokes upfront saves a lot of effort later.  In some 
languages this is the only way to do things, while others allow assigning 
aliases to packages/libraries (e.g., `import pandas as pd; pd.read_stata()`).
That concludes my coding related rant.

```{r, render=FALSE}
# Set the pseudo-random number seed for the sake of replicability
set.seed(7779311)

# Load the library we'll use to parse the Stata formated file
library(haven)

# Load the library we'll use to do some data munging
library(dplyr)

# Storing the file location in a variable for code formatting purposes
filenm <- "https://github.com/OpenSDP/faketucky/raw/master/faketucky.dta"

# Read the dataset into memory, and return only the first 42 variables/columns
df <- haven::read_dta(filenm, 
	  col_select = c("sid", "first_dist_code", "first_hs_code", 
			         "first_hs_alt", "first_hs_urbanicity", "chrt_ninth", 
	  			     "male", "race_ethnicity", "frpl_ever_in_hs", 
	  			     "sped_ever_in_hs", "lep_ever_in_hs", "gifted_ever_in_hs",
	  			     "ever_alt_sch_in_hs", "scale_score_6_math", 
	  			     "scale_score_6_read", "scale_score_8_math", 
	  			     "scale_score_8_read", "pct_absent_in_hs", 
	  			     "pct_excused_in_hs", "avg_gpa_hs", "scale_score_11_eng", 
	  			     "scale_score_11_math", "scale_score_11_read",
	  			     "scale_score_11_comp", "collegeready_ever_in_hs", 
	  			     "careerready_ever_in_hs", "ap_ever_take_class", 
	  			     "last_acadyr_observed", "transferout", "dropout", 
	  			     "still_enrolled", "ontime_grad", "chrt_grad", "hs_diploma",
	  			     "enroll_yr1_any", "enroll_yr1_2yr", "enroll_yr1_4yr",
	  			     "enroll_yr2_any"))

# Rename variables to less verbose but still clear names
names(df) <- c("stdid", "distid", "schcd", "altsch", "urbanicity", 
			   "cohort", "male", "race", "frleverhs", "swdeverhs", "eleverhs",
			   "tageverhs", "alteverhs", "mthss6", "rlass6", "mthss8", 
			   "rlass8", "pctabshs", "pctexcusedhs", "hsgpa", "acteng11", 
			   "actmth11", "actrla11", "actcmp11", "evercollrdyhs", 
			   "evercarrdyhs", "aptakenever", "lastobsyr", "transfer", 
			   "dropout", "stillenrolled", "gradontime", "gradcohort", 
			   "diploma", "yr1psenrany", "yr1psenr2yr", "yr1psenr4yr", 
			   "yr2psenrany")

# Create the unique school identifier by concatenating the dist & sch codes
df$schid <- paste0(df$distid, df$schcd)

# Get a random sample of school IDs
validSchools <- data.frame("schid" = sample(unique(df$schid), size = 60))

# Filter the data set to include only the random sample of school IDs
df <- dplyr::inner_join(df, validSchools)

# New version of R broke the usually way I would do this, but thankfully my 
# spiffy keyboard (SO to Jason Becker Cohort 4) allows me to save time by recording
# and playing back keystrokes so I really only had to type the variable names
df$altsch <- as.factor(df$altsch)
df$cohort <- as.factor(df$cohort)
df$male <- as.factor(df$male)
df$swdeverhs <- as.factor(df$swdeverhs)
df$eleverhs <- as.factor(df$eleverhs)
df$schid <- as.factor(df$schid)
df$tageverhs <- as.factor(df$tageverhs)
df$alteverhs <- as.factor(df$alteverhs)
df$evercollrdyhs <- as.factor(df$evercollrdyhs)
df$evercarrdyhs <- as.factor(df$evercarrdyhs)
df$aptakenever <- as.factor(df$aptakenever)
df$transfer <- as.factor(df$transfer)
df$dropout <- as.factor(df$dropout)
df$stillenrolled <- as.factor(df$stillenrolled)
df$gradontime <- as.factor(df$gradontime)
df$diploma <- as.factor(df$diploma)
df$yr1psenrany <- as.factor(df$yr1psenrany)
df$yr1psenr2yr <- as.factor(df$yr1psenr2yr)
df$yr1psenr4yr <- as.factor(df$yr1psenr4yr)
df$yr2psenrany <- as.factor(df$yr2psenrany)
df$schid <- as.factor(df$schid)
df$race <- as.factor(df$race)
df$urbanicity <- as.factor(df$urbanicity)
df$frleverhs <- as.factor(df$frleverhs)
df$lastobsyr <- as.factor(df$lastobsyr)
df$gradcohort <- as.factor(df$gradcohort)

# Remove the school/district codes (subsumed in schid) and student IDs
df <- df[-c(2, 3)]

# Display a sample of the data
head(df, n = 20)
```


# Simple Synthesis Example
Although this likely is not what you should do in practice, I'll show the simplest 
possible example of generating a synthetic dataset:

```{r}
# Loads the library that is used for synthetic data generation in a decent 
# amount of the SDC literature
library(synthpop)

# Simplest synthetic example:
cake <- synthpop::syn(df)

# Show the synthpop generated object
cake
```

## Why the example above is not generally a good idea
`synthpop` will make a lot of assumptions about your data based on arbitrary 
features, like the order of the variables in the data frame.  In the example,
every variable in the data frame will be used in the synthesis process and they 
will enter from the variable with the lowest index to the highest.  This means 
that the underlying code base is going to start the synthesis process by 
resampling student IDs.  We don't want that to happen though.  So, we need to 
either do some data munging to rearrange the order of the variables in the data
frame or we need to pass some arguments to the function's parameters to make it 
behave the way we want it to behave.

_IMPORTANT_
If you attempt to call the `synthpop::syn()` function with the default method 
and your session freezes or becomes unresponsive, try changing the method to `"ctree"`. 
This is a recommendation from one of the authors of the software and worked for 
the example below (which is why you see `method = "ctree"` in the example below).

```{r}
# Define a new order of columns in the data frame
ord <- c("schid", "altsch", "urbanicity", "male", "race", "cohort", 
		 "frleverhs", "swdeverhs", "eleverhs", "tageverhs", "alteverhs", 
		 "mthss6", "rlass6", "mthss8", "rlass8", "pctabshs", "pctexcusedhs", 
		 "aptakenever", "lastobsyr", "transfer", "dropout", "stillenrolled", 
		 "hsgpa", "gradontime", "gradcohort", "diploma", "evercollrdyhs", 
		 "evercarrdyhs", "actmth11", "actrla11", "acteng11", "actcmp11", 
		 "yr1psenr2yr", "yr1psenr4yr", "yr1psenrany", "yr2psenrany")

# Create the synthetic dataset object, exclude Student IDs, specify the order
# in which the variables are synthesized, and keep the information about the 
# models in the output object.
cakier <- synthpop::syn(df[-c(1)], models = TRUE, method = "ctree", 
						visit.sequence = ord)

# Show the object
cakier
```
# Improvements
The model above exerts some control over the order of the variables being synthesized.
In this example, we resampled the unique school identifiers.  Now, using only the `schid` values from the protected data, we'll *predict* the value of `altsch` 
and use that model to _project_ values of `altsch` in our synthetic data.  
Next, the program will use both the `schid` and `altsch` variables to *predict* 
the `urbanicity` variable and will use that model to _project_ synthetic values 
of `urbanicity` using the resampled `schid` and synthetic `altsch` values.  
This continues one variable at a time until we get to the last variable in the 
`visit.sequence`. 

You can synthesize the data in any order you want, but I selected this order 
based on the assumption that it is going to better reflect the underlying data 
generating process (DGP).  You may have other assumptions and it may not affect 
the end result much at all.

## Things to keep in mind
It isn't discussed on ton in the statistical data control (SDC) literature, but 
one challenge with this approach to synthesis is error propagation. Each model 
has it's own amount of error. That error is injected into the projections.  As 
the process progresses further to the later variables in the data set, the error 
from earlier projections is compounded.  

```{r}
# Now we'll add some logical constraints
cakiest <- synthpop::syn(df[-c(1)], models = TRUE, method = "ctree",
						 visit.sequence = ord,
			rules = list(yr1psenrany = "yr1psenr2yr == 1 | yr1psenr4yr == 1",
						 diploma = "dropout == 0 & stillenrolled == 0 & transfer == 0"),
			  rvalues = list(yr1psenrany = 1, diploma = 1))

# Show the object
cakiest
```
# Further Improved Synthesis 
The improved example has quite a bit more information specified. The `models` 
parameter will save the parameters for the models fitted to the data to generate 
the synthetic records.  However, the last three parameters:
`visit.sequence`, `rules`, and `rvalues` are the most important for controlling 
how the data are synthesized.  `visit.sequence` allows you to specify the order 
in which the variables are synthesized.  This is particularly important since 
the first variable is resampled.  

## Rules and Rule Values
Sometimes variables have logical constraints and we want the synthetic data to 
impose similar logical constraints on the data we synthesize.  The `rules` and 
`rvalues` parameters allow us to specify the logical constraints and values that 
should result from the logical constraints.  In the first example, we have the 
following logical constraint specified as a rule:

`yr1psenrany = "yr1psenr2yr == 1 | yrpsenr4yr == 1"`

This means that the variable `yr1psenrany` is a function of a logical constraint 
for the `yr1psenr2yr` and `yr1psenr4yr` variables.  In otherwords, if a student 
enrolls in the year after completing high school in a 2-year or 4-year 
post-secondary education program, it will affect the `yr1psenrany` value.  When 
the logical constraint is true we specify the value that the `yr1psenrany` 
variable should take in an argument passed to the `rvalues` parameter: 

`yr1psenrany = 1`

So, if the synthetic student enrolls in a 2- or 4-year post-secondary 
institution, we want the synthetic value for the `yr1psenrany` variable to be 
equal to 1, indicating that the synthetic student enrolled in a post-secondary 
program in the year following high school.

# Longitudinal Data
In general, I would recommend using other approaches for longitudinal data synthesis.  
To use the approach here, you'd need to use a multivariate or *wide* data 
structure (e.g., one variable for each measure for each period of time).  This 
increases the error propagation and also requires additional effort to specify 
the order of the variables to maintain the appropriate temporal relationship.
